---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---

## Loading and preprocessing the data
```{r reading}
data <- read.csv("activity.csv")
names(data)
```


## What is mean total number of steps taken per day?
```{r calculatemean}
sdate <- split(data, data$date)
totalstepsbydate <- sapply(sdate, function(x) sum(x[,"steps"]))
hist(totalstepsbydate, main="Histogram of total daily steps", xlab = "total steps by day")
summary(totalstepsbydate, na.rm = TRUE)
mean(totalstepsbydate, na.rm = TRUE)
median(totalstepsbydate, na.rm = TRUE)

```

## What is the average daily activity pattern?
```{r dailypattern}
 sinterval <- split(data, data$interval)
 averagebyinterval<- sapply(sinterval, function(x) mean(x[, "steps"], na.rm = TRUE))
 plot(averagebyinterval, type = "l", main="Average steps by interval", xlab= "5-minute interval", ylab="average steps")
 maxstepsday <- max(averagebyinterval)
 rounded <-round(maxstepsday, digits=2)
 interval <- names(which(averagebyinterval == maxstepsday))
 if(nchar(interval) == 3) time <- paste(substr(interval,1,1), substr(interval, 2, 3), sep = ":")
 if(nchar(interval) == 4) time <- paste(substr(interval,1,2), substr(interval, 3, 4), sep = ":")
 
```
The maximum average number of steps by interval is `r rounded` and it happens in the `r interval` interval, i.e. the interval that starts at `r time`.

## Imputing missing values
```{r missingvalues}
missing <- !complete.cases(data)
totalmissing <- sum(!complete.cases(data))
```
There are `r totalmissing` rows with missing data.

```{r replacingmissingvalues}
## The missing values will be replaced by the average-by-interval values obtained in {r dailypattern}.
newdata <- data
newvalues <- data.frame(averagebyinterval, names(averagebyinterval))
names(newvalues) <- c("average", "name")
names(newvalues)
newdata[missing,]$steps <- if(newdata[missing,]$interval == newvalues$name) newvalues$average
```

```{r calculationswithnewdata}
sdate2 <- split(newdata, newdata$date)
totalstepsbyday <- sapply(sdate2, function(x) sum(x[,"steps"]))
hist(totalstepsbyday, main="Histogram of total daily steps", xlab = "total steps by day")
mean(totalstepsbyday)
median(totalstepsbyday)

```
When replacing missing values by an arbitrary average, the mean and the median become equal.


## Are there differences in activity patterns between weekdays and weekends?
```{r weekdayactivitypatterns}
newdata$weekday <- weekdays(as.Date(newdata$date))
weekday <- c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday")
logicalWeekday <- newdata$weekday %in% weekday
weekdaydata <- newdata[logicalWeekday,]
weekenddata <- newdata[!logicalWeekday,]
s1 <- split(weekdaydata, weekdaydata$interval)
s2 <- split(weekenddata, weekenddata$interval)
weekdaypattern<- sapply(s1, function(x) mean(x[, "steps"]))
weekendpattern<- sapply(s2, function(x) mean(x[, "steps"]))
par(mfrow = c(2,1))
plot(weekdaypattern, type = "l", main="Steps by interval on weekdays", xlab= "5-minute interval", ylab="average steps")
plot(weekendpattern, type = "l", main="Steps by interval on weekends", xlab= "5-minute interval", ylab="average steps")

```

The weekday pattern looks very much like the average pattern, i.e. more steps in the morning, while the weekend pattern shows the steps distributed along the day.
